Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-06-19 paper/main.tex
clean up proofs

For the coming pages, we employ the following notation. Let
\begin{equation}
    \mtx{B}(t) 
    = \mtx{U}(t) \mtx{\Lambda}(t) \mtx{U}(t)^{\top} 
\end{equation}
be the eigenvalue decomposition of $\mtx{B}(t)$ at a fixed $t$, i.e. $\mtx{\Lambda}(t) = \operatorname{diag}(\lambda_1(t), \dots, \lambda_n(t))$ and $\mtx{U}(t)^{\top} \mtx{U}(t) = \mtx{I}_n$. We define the matrix partitions
\begin{equation}
    \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
    \mtx{U}(t) = \begin{bmatrix}
        \smash{\underbrace{\mtx{U}_1(t)}_{n \times k}} & \smash{\underbrace{\mtx{U}_2(t)}_{n \times (n-k)}}
    \end{bmatrix}
    \quad \text{and} \quad
    \mtx{\Lambda}(t) =
    \begin{bmatrix}
        \smash{\overbrace{\mtx{\Lambda}_1(t)}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2(t)}_{(n-k) \times (n-k)}}
    \end{bmatrix}
\end{equation}
and let $\mtx{\Omega}_1(t) = \mtx{U}_1(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2(t) = \mtx{U}_2(t)^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$ with $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$.

\begin{lemma}{Moments of tail bound}{ose}
    Let $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ such that $|\sigma_1| \geq \dots \geq |\sigma_m|$ and independent standard Gaussian random matrices $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all \todo{$k \geq 2$}
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F.
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{proof}
    We can rewrite
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{2p}}
        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
    \end{equation}

    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 2$ as follows
    \begin{equation}
        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}
    \end{equation}
    In case $m$ is not a multiple of $k$, we set $\sigma_i = 0$ if $i > m$. Now we see that
    \begin{equation}
        \sigma_1^2 \leq \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
    \end{equation}
    By the triangle inequality for moments \todo{[cite Hysan]}
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} \notag \\
        &\leq \frac{k + p}{2} \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right)
    \end{align}
    Thus, bounding the constant up front with $1$, we get
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        \leq \sqrt{\frac{k + p}{2}} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F \right)
    \end{equation}
    From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that 
    \begin{equation}
        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq \lVert \mtx{\Sigma} \rVert _2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2 \right] + \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \vct{\omega}_2 \rVert _2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. The second moment is bound by \todo{equation} while for the first moment we apply the above $\mtx{\Sigma} = \mtx{I}_k$ to get
    \begin{equation}
        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F.
    \end{equation}

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \sqrt{2} \cdot \frac{1 + c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
    for a universal constant $c > 0$. In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1})), n_{\mtx{\Omega}} \geq 2$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
\end{lemma}

@@

\begin{proof}
    We let $n_{\mtx{\Omega}} = 2k$ for some $k \in \mathbb{N}$, i.e. $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$. After temporarily omitting the parameter dependence, we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound the integrand with% \todo{maybe explain where $\mtx{\Omega}_i$ come from...}
    \begin{equation}
        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F.
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}

    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term can be bonud with
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
        \label{equ:nystrom-proof-frobenius-trace}
    \end{equation}

    The second term is first processed with standard matrix-norm inequalities:
    \begin{equation}
        \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F 
        \leq \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F
        \leq \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2.
        \label{equ:nystrom-proof-processed-tail}
    \end{equation}
    Its higher order moments are therefore bounded by
    \begin{equation}
        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )^2 \rVert _F \right]
        \leq \sqrt{k} \cdot \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
        = \sqrt{k} \cdot \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2
        \label{equ:nystrom-proof-tail-bound}
    \end{equation}

    We first use the submultiplicativity of the spectral norm and the stochastic independendence of the random matrices
    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
        \leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
        \leq \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right] \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
        \label{equ:OSE-moment-bound-initial-new}
    \end{equation}
    The first term can be bound with \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$:
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq \sqrt{\frac{3}{2}} \cdot \left( 2 \sqrt{k} + \frac{1}{\sqrt{2}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{3}{2}} \cdot \lVert \mtx{\Sigma} \rVert _F.
    \end{equation}
    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
    \begin{equation}
        \mathbb{E}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^{k} \right]
        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{k}{2}} \right]
        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\sfrac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\sfrac{k}{2}}.
    \end{equation}
    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively. Hence,
    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
        \leq \frac{e^2}{k + 1}\sqrt{ \frac{3k}{2} }
        \leq e^2 \sqrt{\frac{3}{2k}}.
        \label{equ:OSE-moment-bound-second}
    \end{equation}

    Putting \refequ{equ:OSE-moment-bound-first} and  \refequ{equ:OSE-moment-bound-second} together, we get
    \begin{equation}
        \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
        \leq ...
    \end{equation}

    Just as in the proof of \cite[lemma 3.1]{meyer-2021-hutch-optimal} we can bound
    \begin{align}
        &\leq \sqrt{k} c \cdot \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2^2 && \text{(\reflem{lem:ose} with $\mtx{A}=\mtx{\Lambda}_2^{\sfrac{1}{2}}$)} \notag \\
        &= \sqrt{k} c \cdot \lambda_{k+1} && \text{(from the definition of $\lVert \cdot \rVert _2$)} \notag \\
        &\leq \frac{c}{\sqrt{k}} \Trace(\mtx{B}) && \text{($\lambda_{k+1} \leq \Trace(\mtx{B})/k$)}
    \end{align}

>>>>

\begin{lemma}{Moment of spectral norm of non-standard Gaussian random matrix}{ose}
    The matrix $\mtx{\Sigma} = \operatorname{diag}(\sigma_1, \dots, \sigma_m) \in \mathbb{R}^{m \times m}$ with $\sigma_1 \geq \dots \geq \sigma_m \geq 0$ and the standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{m \times k}$ satisfy for all $p \in \mathbb{N}$ 
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq  \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
    \end{equation}
    %\begin{equation}
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \mtx{\Omega}_2\rVert _2 \right]
    %    &\leq \sqrt{\frac{k + p}{2}} \cdot \left( 2 + \frac{1}{\sqrt{k}} \right) \cdot \lVert \mtx{\Sigma} \rVert _2 + \sqrt{\frac{k + p}{2k}} \cdot \lVert \mtx{\Sigma} \rVert _F. \notag \\
    %    \mathbb{E}^{p}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]
    %    &\leq \frac{1}{2} \left( 1 + \frac{p}{k - r + 1 - p} \right)^{\sfrac{1}{p}} \left( \frac{k + r}{k - r + 2} \right).
    %\end{equation}
    %If we choose 
\end{lemma}

\begin{proof}
    First, consider the above moment of a vector $\vct{\omega} \in \mathbb{R}^{m}$. We can rewrite
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        = \mathbb{E}\left[  \left( \lVert \mtx{\Sigma} \vct{\omega} \rVert _2^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \mathbb{E}\left[  \left( \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right)^{\sfrac{p}{2}} \right]^{\sfrac{1}{p}}
        = \sqrt{ \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right] }.
        \label{equ:matvec-spectral-norm-simplification}
    \end{equation}
    Inspired by the proof of \cite[theorem 1]{cohen-2016-optimal-approximate}, we split the diagonal entries of $\mtx{\Sigma}$ into $\ell = \lceil m/k \rceil$ groups of size $k \geq 1$
    \begin{equation}
        \underbrace{\sigma_1, \dots, \sigma_k}_{\leq \sigma_1}, \underbrace{\sigma_{k+1}, \dots, \sigma_{2k}}_{\leq \sigma_{k+1}}, \dots, \underbrace{\sigma_{1+(\ell - 1)k}, \dots, \sigma_{\ell k}}_{\leq \sigma_{1+(\ell - 1)k}}.
    \end{equation}
    If $m$ is not a multiple of $k$, we let $\sigma_i = 0$ for $i > m$. Since $\sigma_1 \geq \dots \geq \sigma_{\ell k} \geq 0$,
    \begin{equation}
        \sigma_1^2 = \lVert \mtx{\Sigma} \rVert _2^2
        \quad \text{and} \quad
        \sigma_{1 + (i-1)k}^2 \leq \frac{1}{k} \sum_{j=1}^{k} \sigma_{j + (i-2)k}^2, i = 2, \dots, \ell,
    \end{equation}
    and therefore
    \begin{equation}
        \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \sum_{j=1}^{(\ell - 1)k} \sigma_j^2 \leq \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2.
        \label{equ:singular-value-group-bound}
    \end{equation}
    This allows us to bound
    \begin{align}
        \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{m} \sigma_i^2 \omega_i^2 \right]
        &= \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sum_{j=1}^{k} \underbrace{\sigma_{j + (i-1)k}^2}_{\leq \sigma_{1 + (i-1)k}^2} \omega_{j + (i-1)k}^2 \right] && \text{(separate sum into groups)} \notag \\
        &\leq \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{($\sigma_{j + (i-1)k}^2 \leq \sigma_{1 + (i-1)k}^2$)} \notag \\
        &\leq \sum_{i=1}^{\ell} \sigma_{1 + (i-1)k}^2 \cdot \mathbb{E}^{\sfrac{p}{2}}\left[ \sum_{j=1}^{k} \omega_{j + (i-1)k}^2 \right] && \text{(triangle inequality)} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) 2 \left( \frac{\Gamma(\frac{k + p}{2})}{\Gamma(\frac{k}{2})} \right)^{\sfrac{2}{p}} && \text{(\refequ{equ:singular-value-group-bound} and $\chi_k^2$ moments \cite[theorem 3.3.2]{hogg-2013-introduction-mathematical})} \notag \\
        &\leq \left( \lVert \mtx{\Sigma} \rVert _2^2 + \frac{1}{k} \lVert \mtx{\Sigma} \rVert _F^2 \right) (k + p). && \text{(\reflem{lem:gamma})}
    \end{align}
    Thus, together with \refequ{equ:matvec-spectral-norm-simplification} and using $\sqrt{a+b} \leq \sqrt{a} + \sqrt{b}$ for $a, b \geq 0$, we obtain
    \begin{equation}
        \mathbb{E}^{p}\left[ \lVert \mtx{\Sigma} \vct{\omega} \rVert _2 \right]
        \leq \sqrt{k + p} \left( \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
        \label{equ:matvec-spectral-norm-moment-bound}
    \end{equation}

    Now, we consider the matrix case. From the proof of \cite[lemma B.1]{tropp-2023-randomized-algorithms} we know that 
    \begin{equation}
        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq \lVert \mtx{\Sigma} \rVert _2 \mathbb{E}^p\left[ \lVert \vct{\omega}_1 \rVert _2 \right] + \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \vct{\omega}_2 \rVert _2 \right]
    \end{equation}
    for independent standard Gaussian random vectors $\vct{\omega}_1 \in \mathbb{R}^{k}$ and $\vct{\omega}_2 \in \mathbb{R}^{m}$. Both moments can be bound using \refequ{equ:matvec-spectral-norm-moment-bound}, one of which with $\mtx{\Sigma} = \mtx{I}_k$, to get
    \begin{equation}
        \mathbb{E}^p\left[ \lVert \mtx{\Sigma} \mtx{\Omega} \rVert _2 \right]
        \leq \sqrt{k + p} \left( 3 \lVert \mtx{\Sigma} \rVert _2 + \frac{1}{\sqrt{k}} \lVert \mtx{\Sigma} \rVert _F \right).
    \end{equation}

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuously depend on $t \in [a, b]$. Then its Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ for \emph{even} $n_{\mtx{\Omega}} \in \mathbb{N}$ and $\gamma \geq 1$ verifies with probability at least $1 - \gamma^{-\sfrac{n_{\mtx{\Omega}}}{4}}$
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t.
    \end{equation}
    for a universal constant $c > 0$. In particular, if we choose $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} \log(\delta^{-1}))$, then with probability at least $1-\delta$ for $\delta \in (0, e^{-e^{4 \varepsilon^2}})$ and $\varepsilon > 0$ we have $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$.
\end{lemma}

@@

\begin{proof}
    First, we temporarily omit the parameter dependence. Let $\mtx{B} = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ be the eigenvalue decomposition of a symmetric positive semi-definite matrix $\mtx{B}$, i.e. $\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\mtx{U}^{\top} \mtx{U}= \mtx{I}_n$. We define the matrix partitions
    \begin{equation}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix}
    \end{equation}
    for some \todo{$k \leq n/2$}. Further, for a standard Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times 2k}$ we define $\mtx{\Omega}_1 = \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times 2k}$ and $\mtx{\Omega}_2 = \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times 2k}$, which are likewise standard Gaussian.

    We will set $n_{\mtx{\Omega}} = 2k$, such that we can apply \cite[theorem B.1]{persson-2023-randomized-lowrank} with $f(x) = x$ to bound
    \begin{equation}
        \lVert \mtx{B} - \Nystr{\mtx{\Omega}}{\mtx{B}} \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm.

    As in the proof of \cite[lemma 3]{meyer-2021-hutch-optimal}, the first term is limited by
        \lVert \mtx{\Lambda}_2 \rVert _F
        = \sqrt{\sum_{i=k+1}^{n} \lambda_i^2}
        \leq \sqrt{ \lambda_{k+1} \sum_{i=k+1}^{n} \lambda_i}
        \leq \sqrt{ \frac{\Trace(\mtx{B})}{k} \Trace(\mtx{B})}
        \leq \frac{1}{\sqrt{k}} \Trace(\mtx{B}).
        \label{equ:nystrom-proof-frobenius-trace}
    The second term is first processed with standard matrix norm manipulations to
    \begin{align}
        \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]
        &= \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert (\mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} )( \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger})^{\top} \rVert _F \right] && \text{($\lVert \mtx{A} \rVert _{(4)}^2 = \lVert \mtx{A} \mtx{A}^{\top} \rVert _{F}$)} \notag \\
        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _F \right] && \text{($\lVert \mtx{A} \mtx{B} \rVert _F \leq \lVert \mtx{A} \rVert _F \lVert \mtx{B} \rVert _2$)} \notag \\
        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{($\lVert \mtx{A} \rVert _F \leq \sqrt{k} \lVert \mtx{A} \rVert _2$)} \notag \\
        &\leq \mathbb{E}^{\sfrac{k}{2}}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2 \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]  && \text{(submultiplicativity)} \notag \\
        &\leq \sqrt{k} \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{\sfrac{k}{2}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right]  && \text{(independence $\mtx{\Omega}_1, \mtx{\Omega}_2$)} \notag \\
        &= \sqrt{k} \mathbb{E}^{k}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]^2 \mathbb{E}^{k}\left[  \lVert \mtx{\Omega}_1^{\dagger} \rVert _2 \right]^2 && \text{(definition of $\mathbb{E}^{k}[\cdot]$)}
        \label{equ:nystrom-proof-processed-tail}
    \end{align}
    We can apply \reflem{lem:ose} with $\mtx{\Sigma} = \mtx{\Lambda}_2^{\sfrac{1}{2}}$ to bound the first term with
        \mathbb{E}^k\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2 \right]
        \leq \sqrt{3 k} \left( 3 \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _2 + \frac{1}{\sqrt{2k}} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \rVert _F \right).
        \label{equ:spectral-norm-bound-applied}
    The moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ can be bound with help of the proof of \cite[lemma B.3]{tropp-2023-randomized-algorithms}
        \mathbb{E}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^{k} \right]
        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{k}{2}} \right]
        \leq \left( k + 1 \right) \left( \frac{1}{(k + 1)!}\right)^{\sfrac{k}{k + 1}} \left( \frac{3 k}{2}\right)^{\sfrac{k}{2}}.
    With the Taylor series expansion of the exponential function it can be shown that $e^n \geq 1 + n$ and $e^n \geq \frac{n^n}{n!}$ for all $n \in \mathbb{N}$, from which $(n + 1)^{\sfrac{1}{n}} \leq e$ and $\left( \frac{1}{n!} \right)^{\sfrac{1}{n}} \leq \frac{e}{n}$ follow respectively. Hence,
